---
title: "Resampling example"
author: "Joey Burant"
date: "1/29/2022"
output: html_document
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE)

## required packages
library(tidyverse)
library(palmerpenguins)
library(patchwork)

```

This is a quick example of how to resample the data and plot the distributions of sampled traits, making use of the `penguins` dataset
from the `palmerpenguins` **R** package.

### Resampling the data using `dplyr::sample_()` and `base::replicate()`

We can easily randomly resample either a specific number of observations (using `dplyr::sample_n()`) or a set proportion of the data (using `dplyr::sample_frac()`). The `replicate()` function provides a really easy way to iterate this process a number of times.

The code chunk below uses two variables:  

- `n_reps` = the number of desired replicates, which is passed to `replicate()` -- in this case we want 100 replicates. 
- `samp_prop` = the proportion of the data to resample, which is passed to `sample_frac()` -- here we want to resample 50% of the data

By wrapping this in a tibble, we can create a tidy product in which the replicate samples are stored as a nested list that can be used for summarizing, analysis, and plotting.

```{r resampling}

## remove NAs from the dataset and add a row (record) id column
## (so we can trace a sample back to the dataset if needed)
penguins_c <- penguins %>% 
  drop_na() %>% 
  mutate(id = row_number(), 
         .before = species)

## specify the number of replicates and the sample prop (or size)
n_reps <- 100
samp_prop <- 0.50

## resample the data and put it in a tibble
re_penguins <- tibble(
  rep = 1:n_reps, 
  samp = replicate(
    n = n_reps, 
    sample_frac(
      tbl = penguins_c, 
      size = samp_prop, 
      replace = FALSE), 
    simplify = FALSE))

```

What does this look like now?

```{r look nested table}

re_penguins

```


### Summarizing the data

One way to summarize the dataset is to first `unnest()` it, then use the replicate identifier (`rep`) as a grouping variable. 

In the chunk below, I first `unnest()` the data (stored in the `samp` column), then `group_by()` `rep` (100 levels), `species` (3 levels), and `sex` (2 levels). This example calculates the median value for each of the numeric columns in the dataset, before `nest`ing the summary data and `join`ing it back with the original data.

```{r summarizing nested data}

## calculate the median trait values from each sample
re_penguins <- re_penguins %>% 
  unnest(samp) %>% 
  group_by(rep, species, sex) %>% 
  summarise(across(.cols = bill_length_mm:body_mass_g, 
                   ~ median(.))) %>% 
  group_by(rep) %>% 
  nest() %>% 
  rename(summ_mean = data) %>% 
  right_join(re_penguins, ., by = "rep") %>% 
  suppressMessages() 

re_penguins

```

### Plotting

Finally, we can plot the distributions of the re-sampled data. The plot below overlays density distributions of `penguins$bill_length_mm` for each random sample of the dataset, facetting by `species` and `sex`.

```{r plotting the distribution of bill length}

## plot the distributions of bill_length_mm from the resamples
re_penguins %>% 
  select(rep, samp) %>% 
  unnest(samp) %>% 
  ggplot(data = ., 
         mapping = aes(x = bill_length_mm, 
                       group = rep, 
                       colour = sex)) + 
  geom_line(stat = "density", alpha = 0.2) + 
  scale_colour_brewer(palette = "Dark2") + 
  # facet_wrap(~ species) + 
  facet_grid(species ~ sex) + 
  labs(x = "Bill length (mm)", y = "Density", 
       title = "Distributions of bill length in resamples of the penguins dataset", 
       subtitle = "50% of the data were randomly resampled 100 times", 
       caption = "Source: palmerpenguins R package") + 
  theme_minimal() + 
  theme(legend.position = "none")

```

This plot provides a good starting point for thinking about how to calculate a 95% confidence interval on the distributions based on the distribution of resamples. Of course, the same approach could also be used to estimate confidence intervals on the mean/median difference between two distributions (e.g., males and females, or comparing species).

#### What about the distributions of all the traits?

```{r plot all distributions, echo = FALSE}

## use patchwork to build the multi-plot
## bill length
p1.bl <- re_penguins %>% 
  select(rep, samp) %>% 
  unnest(samp) %>% 
  ggplot(data = ., 
             mapping = aes(x = bill_length_mm, 
                           group = interaction(rep, sex))) + 
  geom_line(mapping = aes(colour = sex),
            stat = "density", alpha = 0.2) +
  # geom_histogram(mapping = aes(fill = sex), 
  #                alpha = 0.5, position = "identity") + 
  scale_colour_brewer(palette = "Dark2", 
                      name = "", labels = c("female", "male")) + 
  # facet_grid(species ~ measure, 
  #            scales = "free") + 
  facet_wrap(~ species, ncol = 1) + 
  labs(x = NULL, y = "Density", 
       subtitle = "Bill length (mm)") + 
  theme_minimal() + 
  theme(legend.position = "none", 
        plot.subtitle = element_text(hjust = 0.5, face = "italic"), 
        strip.text = element_blank(), 
        strip.background = element_blank())
# p1.bl

## bill depth
p2.bd <- re_penguins %>% 
  select(rep, samp) %>% 
  unnest(samp) %>% 
  ggplot(data = ., 
         mapping = aes(x = bill_depth_mm, 
                       group = interaction(rep, sex))) + 
  geom_line(mapping = aes(colour = sex),
            stat = "density", alpha = 0.2) +
  # geom_histogram(mapping = aes(fill = sex), 
  #                alpha = 0.5, position = "identity") + 
  scale_colour_brewer(palette = "Dark2", 
                      name = "", labels = c("female", "male")) + 
  # facet_grid(species ~ measure, 
  #            scales = "free") + 
  facet_wrap(~ species, ncol = 1) + 
  labs(x = NULL, y = NULL, 
       subtitle = "Bill depth (mm)") + 
  theme_minimal() + 
  theme(legend.position = "none", 
        plot.subtitle = element_text(hjust = 0.5, face = "italic"), 
        strip.text = element_blank(), 
        strip.background = element_blank())
# p1.bd

## flipper length
p3.fl <- re_penguins %>% 
  select(rep, samp) %>% 
  unnest(samp) %>% 
  ggplot(data = ., 
         mapping = aes(x = flipper_length_mm / 10, 
                       group = interaction(rep, sex))) + 
  geom_line(mapping = aes(colour = sex),
            stat = "density", alpha = 0.2) +
  # geom_histogram(mapping = aes(fill = sex), 
  #                alpha = 0.5, position = "identity") + 
  scale_colour_brewer(palette = "Dark2", 
                      name = "", labels = c("female", "male")) + 
  # facet_grid(species ~ measure, 
  #            scales = "free") + 
  facet_wrap(~ species, ncol = 1) + 
  labs(x = NULL, y = NULL, 
       subtitle = "Flipper length (cm)") + 
  theme_minimal() + 
  theme(legend.position = "none", 
        plot.subtitle = element_text(hjust = 0.5, face = "italic"), 
        strip.text = element_blank(), 
        strip.background = element_blank())
# p3.fl

## body mass
p4.bm <- re_penguins %>% 
  select(rep, samp) %>% 
  unnest(samp) %>% 
  ggplot(data = ., 
         mapping = aes(x = body_mass_g / 1000, 
                       group = interaction(rep, sex))) + 
  geom_line(mapping = aes(colour = sex),
            stat = "density", alpha = 0.2) +
  # geom_histogram(mapping = aes(fill = sex), 
  #                alpha = 0.5, position = "identity") + 
  scale_colour_brewer(palette = "Dark2", 
                      name = "", 
                      labels = c("female", "male"), 
                      guide = guide_legend(override.aes = list(alpha = 1))) + 
  # facet_grid(species ~ measure, 
  #            scales = "free") + 
  facet_wrap(~ species, ncol = 1, strip.position = "right") + 
  labs(x = NULL, y = NULL, 
       subtitle = "Body mass (kg)") + 
  theme_minimal() + 
  theme(legend.position = "bottom", 
        legend.justification = "right", 
        plot.subtitle = element_text(hjust = 0.5, face = "italic"), 
        strip.text = element_text(face = "bold", size = 12))
# p4.bm

p1.bl + p2.bd + p3.fl + p4.bm + 
  plot_layout(ncol = 4) + 
  plot_annotation(
    title = "Distributions of trait measurements in resamples of the penguins dataset", 
    subtitle = "50% of the observations were randomly resampled 100 times", 
    caption = "Source: palmerpenguins R package")

ggsave("trait_distributions_resamp.pdf", 
       plot = last_plot(), width = 11, height = 8, units = "in", dpi = "retina")

```

**Note**: originally tried building this plot buy pivoting the data to long format at facetting by species and measure (bill length, etc.), but `facet_grid()` doesn't allow for free y axes on each pane so density doesn't scale properly...). There's a workaround with `facet_wrap(~ a*b)` but that creates bad labels and makes it hard to compare distributions across species.
