---
title: "Resampling example"
author: "Joey Burant"
date: "1/29/2022"
output: html_document
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE)

## required packages
library(tidyverse)
library(palmerpenguins)

## import and clean data
## (remove NAs from the dataset)
penguins_c <- palmerpenguins::penguins %>% drop_na()

```

This is a quick example of how to resample the data and plot the distributions of sampled traits, making use of the `penguins` dataset
from the `palmerpenguins` **R** package.

### Resampling the data using `dplyr::sample_()` and `base::replicate()`

We can easily randomly resample either a specific number of observations (using `dplyr::sample_n()`) or a set proportion of the data (using `dplyr::sample_frac()`). The `replicate()` function provides a really easy way to iterate this process a number of times.

The code chunk below uses two variables:  

- `n_reps` = the number of desired replicates, which is passed to `replicate()` -- in this case we want 100 replicates. 
- `samp_prop` = the proportion of the data to resample, which is passed to `sample_frac()` -- here we want to resample 50% of the data

By wrapping this in a tibble, we can create a tidy product in which the replicate samples are stored as a nested list that can be used for summarizing, analysis, and plotting.

```{r resampling}

## specify the number of replicates and the sample prop (or size)
n_reps <- 100
samp_prop <- 0.50

## resample the data and put it in a tibble
re_penguins <- tibble(
  rep = 1:n_reps, 
  samp = replicate(
    n = n_reps, 
    sample_frac(
      tbl = penguins_c, 
      size = samp_prop, 
      replace = FALSE), 
    simplify = FALSE))

```

What does look like now?

```{r look nested table}

re_penguins

```


### Summarizing the data

One easy way to summarize the dataset is to first `unnest()` it, then use the replicate identifier (`rep`) as a grouping variable. 

In the chunk below, I first `unnest()` the data (stored in the `samp` column), then `group_by()` `rep` (100 levels), `species` (3 levels), and `sex` (2 levels). This example calculates the median value for each of the numeric columns in the dataset, before `nest`ing the summary data and `join`ing it back with the original data.

```{r summarizing nested data}

## calculate the median trait values from each sample
re_penguins <- re_penguins %>% 
  unnest(samp) %>% 
  group_by(rep, species, sex) %>% 
  summarise(across(.cols = bill_length_mm:body_mass_g, 
                   ~ median(.))) %>% 
  group_by(rep) %>% 
  nest() %>% 
  rename(summ_mean = data) %>% 
  right_join(re_penguins, ., by = "rep") %>% 
  suppressMessages() 

re_penguins

```

### Plotting

Finally, we can easily plot the distributions of the re-sampled data. The plot below overlays density distributions of `penguins$bill_length_mm` for each random sample of the dataset, facetting by `species` and `sex`.

```{r plotting the distribution of bill length}

## plot the distributions of bill_length_mm from the resamples
re_penguins %>% 
  select(rep, samp) %>% 
  unnest(samp) %>% 
  ggplot(data = ., 
         mapping = aes(x = bill_length_mm, 
                       group = rep, 
                       colour = sex)) + 
  geom_line(stat = "density", alpha = 0.2) + 
  scale_colour_brewer(palette = "Dark2") + 
  # facet_wrap(~ species) + 
  facet_grid(species ~ sex) + 
  labs(x = "Bill length (mm)", y = "Density", 
       title = "Distributions of bill length in resamples of the penguins dataset", 
       subtitle = "50% of the data were randomly resampled 100 times", 
       caption = "Source: palmerpenguins R package") + 
  theme_minimal() + 
  theme(legend.position = "none")

```

This plot would provide a good starting point to start thinking about how to calculate a 95% confidence interval on the distributions based on the distribution of resamples. Of course, the same approach could also be used to estimate confidence intervals on the mean/median difference between two distributions (e.g., males and females, or comparing species).

#### What about the distribution of all the traits?

```{r plot all distributions}

## plot the distributions of all the traits from the resamples
re_penguins %>% 
  select(rep, samp) %>% 
  unnest(samp) %>% 
  pivot_longer(., cols = bill_length_mm:body_mass_g, 
               names_to = "measure", values_to = "value") %>% 
  mutate(measure = factor(.$measure, 
                          levels = c("bill_depth_mm", 
                                     "bill_length_mm", 
                                     "flipper_length_mm", 
                                     "body_mass_g"), 
                          labels = c("Bill depth (mm)", 
                                     "Bill length (mm)", 
                                     "Flipper length(mm)", 
                                     "Body mass (g)"))) %>% 
  # mutate(value = ifelse(measure == "Body mass (g)", value / 1000, value)) %>% 
  ggplot(data = ., 
         mapping = aes(x = value, 
                       group = interaction(rep, sex))) + 
  geom_line(mapping = aes(colour = sex),
            stat = "density", alpha = 0.2) +
  # geom_histogram(mapping = aes(fill = sex), 
  #                alpha = 0.5, position = "identity") + 
  scale_colour_brewer(palette = "Dark2", 
                      name = "", labels = c("female", "male")) + 
  # facet_grid(species ~ measure, 
  #            scales = "free") + 
  facet_wrap(~ species*measure, ncol = 4, scales = "free") + 
  labs(x = NULL, y = "Density", 
       title = "Distributions of trait measurements in resamples of the penguins dataset", 
       subtitle = "50% of the data were randomly resampled 100 times", 
       caption = "Source: palmerpenguins R package") + 
  theme_minimal() + 
  theme(legend.position = "bottom")

```

**Note**: this plot needs some work (facet labels are ugly, but `facet_grid()` doesn't allow for free y axes on each pane so density doesn't scale properly...)
